Capabilities, MMIO and Robust Safety

### 2020-04-03 capability-notes/driver-notes/main.tex

    \X{bootCont_{synt}}(P, g, b, e, A_{\X{entry}},\Phi) \eqdef  \\[0.2em]
    \begin{array}{l}
       (\exists b_{adv}\; e_{adv}\; t.\; \\[0.5em]
      \quad
      \begin{array}{l}
        \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, b_{adv}, e_{adv}, b_{adv}) \ast {} \\[0.7em]
        \bigast{a_{i} \in A_{\X{entry}}}{r_{a_{i}} \mapsto_{r} (\X{E},g,b,e,a_i)} \ast
        \bigast{r \in Reg\setminus \{\X{PC}, A_{\X{entry}}\}}{r \mapsto_{r} \X{inl}\; \_} \ast {} \\[1.4em]
        \bigast{m \in [b_{adv}, e_{adv})}{m \mapsto_{a} \X{inl}\; \_} \ast
        %\bigast{m \in [b, e)}{m \mapsto_{a} \_} \ast
        \driverC(b,e,\overline{\V{w}_l})\ast
        \; \tracefrag{t} \ast \pure{P(t)})
      \end{array}
    \end{array} \\[4em]
    \quad\wand
    \begin{array}{l}
     \wpre{\I{Seq}\; (\I{Instr}\; \I{Executable})}{\Phi}
    \end{array}\\[1.2em]

@@

 \X{bootCont_{sem}}(P, g, b, e, A_{\X{entry}},\Phi,\Winit) \eqdef  \\[0.2em]
    \begin{array}{l}
       (\exists b_{\X{adv}}\; e_{\X{adv}}\; t.\; \\[0.5em]
      \quad
      \begin{array}{l}
        \X{InitW}(\Winit,b_{\X{adv}}, e_{\X{adv}})\ast \\[0.5em]
        \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, b_{\X{PC}}, e_{\X{PC}}, a_{\X{PC}})
        \ast \VR^{\notMMIO} (\Winit)(\X{RW}, \X{G}, b_{\X{PC}}, e_{\X{PC}}, a_{\X{PC}})) \ast {} \\[1.2em]
        \bigast{r \in Reg\setminus \{\X{PC}\}}{(r \mapsto_{r} w_r \ast
        \VR^{\notMMIO} (\Winit)(w_r))} \ast {} \\[1.2em]
        \driverC(b,e,\overline{\V{w}_l})\ast
        \; \tracefrag{t} \ast  \pure{P(t)}\ast \\[0.5em]
        \X{region}(\Winit) \ast  \X{sts\_full\_world}\; \X{sts\_std} \;
        \Winit)\\[0.5em]       \end{array}
    \end{array} \\[4em]
    \quad\wand
    \begin{array}{l}
     \wpre{\I{Seq}\; (\I{Instr}\; \I{Executable})}{\Phi}
    \end{array}\\[1.2em]

@@

>>>>

    \X{bootCont_{synt}}(P, g, b, e, A_{\X{entry}},\Phi) \eqdef  \\[0.2em]
    \begin{array}{l}
       (\exists b_{adv}\; e_{adv}\; t.\; \\[0.5em]
      \quad
      \begin{array}{l}
        \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, b_{adv}, e_{adv}, b_{adv}) \ast {} \\[0.7em]
        \bigast{a_{i} \in A_{\X{entry}}}{r_{a_{i}} \mapsto_{r} (\X{E},g,b,e,a_i)} \ast
        \bigast{r \in Reg\setminus \{\X{PC}, A_{\X{entry}}\}}{r \mapsto_{r} \X{inl}\; \_} \ast {} \\[1.4em]
        \bigast{m \in [b_{adv}, e_{adv})}{m \mapsto_{a} \X{inl}\; \_} \ast
        \bigast{l \in [b, e)}{m \mapsto_{a}\V{w}_l} \ast
        % \driverC(b,e,\overline{\V{w}_l})\ast
        \; \tracefrag{t} \ast \pure{P(t)})
      \end{array}
    \end{array} \\[4em]
    \quad\wand
    \begin{array}{l}
     \wpre{\I{Seq}\; (\I{Instr}\; \I{Executable})}{\Phi}
    \end{array}\\[1.2em]

@@

 \X{bootCont_{sem}}(P, g, b, e, A_{\X{entry}},\Phi,\Winit) \eqdef  \\[0.2em]
    \begin{array}{l}
       (\exists b_{\X{adv}}\; e_{\X{adv}}.\; \\[0.5em]
      \quad
      \begin{array}{l}
        \X{InitW}(\Winit,b_{\X{adv}}, e_{\X{adv}})\ast \\[0.5em]
        \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, b_{\X{PC}}, e_{\X{PC}}, a_{\X{PC}})
        \ast \VR^{\notMMIO} (\Winit)(\X{RW}, \X{G}, b_{\X{PC}}, e_{\X{PC}}, a_{\X{PC}})) \ast {} \\[1.2em]
        \bigast{r \in Reg\setminus \{\X{PC}\}}{(r \mapsto_{r} w_r \ast
        \VR^{\notMMIO} (\Winit)(w_r))} \ast {} \\[1.2em]
        \driverC(b,e,\overline{\V{w}_l})\ast
        \; \trInv \; \ast \\[0.5em]
        \X{region}(\Winit) \ast  \X{sts\_full\_world}\; \X{sts\_std} \;
        \Winit)\\[0.5em]       \end{array}
    \end{array} \\[4em]
    \quad\wand
    \begin{array}{l}
     \wpre{\I{Seq}\; (\I{Instr}\; \I{Executable})}{\Phi}
    \end{array}\\[1.2em]

@@

\subsection{Adapting WP and the FTLR}
The $\I{Load}$ and $\I{Store}$ cases will require new WP rules, that do not mention
points-to chunks but mention the trace resource instead (and specifically, we
will need rules in the presence of invariants such as $\trInv$ above ), when interactions at
\MMIO locations happen. The general WP spec will need to distinguish between
the MMIO and non-MMIO cases (alternatively, if both cases are too distinct, the
ftlr can make this distinction again).

The relations $\VR,\ER,\RR$ need to be updated in terms of $\notMMIO$ and the FTLR will need to be re-stated in terms of these updated relations. This should again only
require some changes for the $\I{Load}$ and $\I{Store}$ cases, depending on how
we choose to formalize the WP cases above.

\subsection{Proving the driver safe}
Specifically, we need to prove the condition

\[
\forall a_{i} \in A_{\X{entry}} \ldotp\,\left\{
    {\begin{array}{ll}
    \trInv \ast
    \driverC(b,e,\overline{\V{w}_l})\!
         \vdash \\
         \forall W \Wpriv \Winit \ldotp\;\ER^{\notMMIO} (W) (\X{RX}, g, b, e, a_i)
        \end{array}} \right.
\]

mentioned above, for the set $A_{\X{entry}}$ of entry points that the boot code
will actually hand closures out for. This will require reductive reasoning over the concrete code of
the driver, but should not be too hard in the absence of local return pointers
and local driver state.

\subsection{Syntactic boot spec}

Proving the syntactic boot specification given above, again through reasoning on
the concrete boot code that we have written down in our example. This should be
relatively easy, as our boot code erases adversary memory before handing over
control. The proof includes verifying any IO that the boot code might need to
do, using the new WP rules mentioned above.

\subsection{Semantic boot spec}

The semantic boot specification should be derivable from the syntactic one,
combined with the proof of safety of the driver (so that we know that enter
capabilities to the driver are safe). The different elements of the semantic bootSpec continuation can be proven as follows:
\begin{itemize}
\item First off, we can prove \[\exists \overline{\gamma}\ldotp \X{region}(\emptyset) \ast  \X{sts\_full\_world}\;
  \X{sts\_std} \; \emptyset\] for \emph{some} ghost names $\overline{\gamma}$ (we
  left this existential out for simplicity in our spec). This will
  suffice, since all our other proofs are parametric in the concrete ghost names, and we
  can use these specific ghost names to construct an instance of $\X{STSG}$. We
  can then populate the world with a standard, permanent region for the
  adversary's memory, using the chunks $\bigast{m \in [b_{adv}, e_{adv})}{m
    \mapsto_{a} \X{inl}\; \_}$ from the syntactic boot spec. This allows us to
  prove \[\exists \overline{\gamma}\ldotp \X{region}(\Winit) \ast  \X{sts\_full\_world}\;
    \X{sts\_std} \; \Winit\] in the semantic bootspec.
  We could generalize the semantic bootspec slightly, by only requiring $[b_{adv},
  e_{adv})$ to be a subset of adversary memory, but it is not immediately clear
  what we gain from that.
\item $\trInv$ and $\driverC(b,e,\overline{\V{w}_l})$ are trivially proven by
  allocating invariants for $\tracefrag{t} \ast \pure{P(t)}$ and $\bigast{l \in [b, e)}{m \mapsto_{a}\V{w}_l}$ respectively.
\item In proving that
  \[\bigast{r \in Reg\setminus \{\X{PC}\}}{(r \mapsto_{r} w_r \ast
      \VR^{\notMMIO} (\Winit)(w_r))}\]
  holds, we can distinguish two different cases:
  \begin{itemize}
  \item For registers $a_i$ containing closures to the driver, we use the chunks
    $\bigast{a_{i} \in A_{\X{entry}}}{r_{a_{i}} \mapsto_{r} (\X{E},g,b,e,a_i)}$
    from the syntactic bootspec, together with the proof of driver safety. To be
    able to apply the latter, we have to use the $\driverC(b,e,\overline{\V{w}_l})$ and $\trInv$ invariants we allocated earlier.
  \item All other registers will contain integers $\X{inl}\; z$ and will hence
    trivially be in $\VR$. We can just use the chunks $\bigast{r \in
      Reg\setminus \{\X{PC}, A_{\X{entry}}\}}{r \mapsto_{r} \X{inl}\; \_} $ to
    satisfy the proof obligation.
  \end{itemize}
\item The proof of
  \[\X{PC} \mapsto_{r} (\X{RWX}, \X{G}, b_{\X{adv}}, e_{\X{adv}}, b_{\X{adv}})
        \ast \VR^{\notMMIO} (\Winit)({\color{red}\X{RW}}, \X{G}, b_{\X{adv}}, e_{\X{adv}}, a_{\X{adv}}))
  \]
   follows directly from $ \X{PC} \mapsto_{r} (\X{RWX}, \X{G}, b_{adv}, e_{adv},
   b_{adv}) $ in the syntactic boot spec, combined with the shape of $\Winit$
   (i.e. the PC points to a standard, permanent region).
\item The execution condition for the PC with respect to $\Winit$ can be proven
  using the fundamental theorem, since we know from the previous bullet that
  both the read- and write-conditions must be satisfied for the PC register.
  This completes the proof of $\VR^{\notMMIO} (\Winit)({\color{red}\X{RWX}}, \X{G}, b_{\X{adv}}, e_{\X{adv}}, a_{\X{adv}}))
$.
\end{itemize}

\subsection{Proving the robust safety theorem}
Given a proof of the semantic boot spec, it should be fairly simple to
prove that the robust safety theorem holds.

To prove that our entire execution is safe, we will first of all need to prove
that $\X{bootCont_{sem}}(P, g, b, e, A_{\X{entry}},\Phi,\Winit)$ holds. This
should be simple, since part of the precondition is that the PC is in the
execution relation, which is essentially what we need to prove (under sensible
-i.e.\ safe-
memory and registers).

When we know that our entire execution is safe according to the WP predicate, we
can apply adequacy of the weakest precondition (and the fact that we have an
invariant stating that our predicate $P$ must hold), to prove $P(t)$ over
the entire I/O trace $t$.
